Machine Learning Models for the Segmentation of eCoaching Text

### 2018-03-05 paper-amia-2018/manuscript.tex
finished first draft without proofread

\noindent{\bf Abstract}
\textit{Poor eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern among young adults. An effective intervention is required to improve eating habits. eCoaching is an email-based intervention technique where a critical step is the segmentation of text for the automatic annotation of the email exchange. In this study, we transformed this task into the classification of detecting the boundary of segmentation and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-score in overall, 0.779 and 0.993 F1-scores for detecting boundary, not a boundary, respectively. This study has a great implication to identify individual text segments, which can be annotated directly with a classification model, and accelerate the pace of identifying effective communication strategies linked to healthy eating.}

@@

\subsection*{\textit{Evaluation metrics}}
In this experiment, standard metrics: precision, recall, and F-measure, are applied to evaluate the performance of binary classifiers\cite{aas1999text}. However, we didn't report accuracy as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. We conduct the experiment with 5 folds cross-validation and weighted macro-averaging of these metrics over the folds. All models have trained on 80\% of the word pairs and remaining 20\% of the data is used as a test set for reporting the performance of the model. We also estimated the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}.

\section*{Results}
Experimental results are evaluated with weighted average, ``boundary'', and ``not boudary'' classes shown in Table~\ref{tab:result_boundary}, \ref{tab:result_not_boundary} and \ref{tab:result_weighted_avg}, respectively.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN for detecting boundaries of segmentation in eCoaching text. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

As follows from Table~\ref{tab:result_boundary}, KNN performs best among all machine learning models in terms of precision and F1-score, achieved 0.808 and 0.728 when contextual features are used, and 0.820 and 0.779 when a combination of contextual topic, and punctuation mark features are used. However, RNN demonstrates the lowest performance among all models in terms of recall and F1-score while GRU shows 3.72\%, 17.79\% and 11.47\% higher precision, recall, and F1-score than LSTM. SVM is the second highest model in terms of precision and F1-score and demonstrate the highest model by achieving 0.679 recall when only textual features are used. On the other hand, NB exhibits lowest precision value 0.594,  among all models but performs better than RNN with recall and F1-score.
When additional features combined with textual features, recall increases 0.6\%, 2.5\%, and 11.92\%, and F1-score increases 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively. Nevertheless, precision increases 4.31\% and 1.49\% for SVM and KNN models, except NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN  for detecting no boundaries in eCoaching text. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

Table~\ref{tab:result_not_boundary} illustrates the performance of NB, SVM, KNN, and RNN models for detecting no boundaries in eCoaching text. Similar to boundary detection, KNN consistently outperformed over all other models and RNN displays the lowest performance in terms of all performance metrics for 'not boundary' identification. Moreover, performance is very high in ``not boundary'' class compared to boundary detection. Effect of the additional feature is also consistent with 'not boundary' classification. Results show that precision, recall, and F1-score increases x\%, y\%, and z\%, for NB, SVM, KNN models, respectively. \\

\begin{table}[ht]
\centering
\caption{Weighted average performance of NB, SVM, KNN, and RNN for the segmentation of eCoaching text in detecting both ``boundary'' and ``not boundary''. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} depicts the performance of NB, SVM, KNN, and RNN models for the segmentation of eCoaching text in detecting both ``boundary'' and ``not boundary'' classes. Overall, KNN performed best and RNN is the lowest among all models for 'not boundary' identification. SVM and RNN achieved x\% and xx\%, y\% and yy\%, z\% and zz\% for precision, recall, and F1, respectively. Effect of the additional feature is also consistent as above, precision, recall, and F1 increases x\%, y\%, and z\%, for NB, SVM, and KNN, respectively.\\

\section*{Discussion}
This study is the first large-scale efforts to evaluate the segmentation of eCoaching text. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-score in overall, 0.779 and 0.993 F1-scores for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn information from the email exchange. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the ecoacing domain, and it can be successfully applied to other domain as well.

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-score were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which was demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values and achieved 0.978 for both cases while it provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively.

@@

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. Similarly, additional information is the common reason for classified original segment into multiple segments. For example, [need help from April].

@@

As for future work, we plan to evaluate our methods with several similar and different datasets. We also plan to use a combination of machine learning and natural language approaches to improve model performance. For example, part-of-speech tagging and distance from the boundary of the sentence might significantly enhance the classification performance. The limitation of this study is that eCoaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings.

\section*{Conclusion}
Segmentation of eCoaching text is an integral part of developing an automated eCoaching intervention. Although several studies have done for the segmentation of clinical text into sections and sentences, none of them are used for the segmentation of text into a group of MI behavior in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop effective behavioral interventions. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model and increase the effectiveness of behavioral interventions. This approach will help for developing fully automated eCoaching and also accelerate the pace of identifying effective communication strategies linked to healthy eating. As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis for enhancing our proposed method.

>>>>

\noindent{\bf Abstract}
\textit{Poor eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern among young adults. An effective intervention is required to improve eating habits. eCoaching is an email-based intervention technique where a critical step is the segmentation of text for the automatic annotation of the email exchange. In this study, we transformed this task into the classification of detecting the boundary of segmentation and developed several state-of-the-art machine learning models including Support Vector Machine, Naive Bayes, K-Nearest Neighbor (KNN), Recurrent Neural Networks by utilizing contextual, topic and punctuation mark features. Results indicate that KNN is the best model and achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. This study has a great implication to identify individual text segments, which can be annotated directly with a classification model, and accelerate the pace of identifying effective communication strategies linked to healthy eating.}

@@

\subsection*{\textit{Evaluation metrics}}
In this experiment, standard metrics: precision, recall, and F1-measure, are applied to evaluate the performance of binary classifiers\cite{aas1999text}. However, we didn't report accuracy as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. We conduct the experiment with 5 folds cross-validation and weighted macro-averaging of these metrics over the folds. All models have trained on 80\% of the word pairs and remaining 20\% of the data is used as a test set for reporting the performance of the model. We also estimated the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}.

\section*{Results}
Experimental results are evaluated with ``boundary'' and ``not boundary'' classes as well as their weighted average, which are shown in Table~\ref{tab:result_boundary}, \ref{tab:result_not_boundary} and \ref{tab:result_weighted_avg}, respectively.\\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN for detecting boundaries of segmentation in eCoaching text. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

As follows from Table~\ref{tab:result_boundary}, KNN performs best among all machine learning models in terms of precision and F1-measure, achieved 0.808 precision with 0.728 F1-measure when contextual features are used, and 0.820 precision with 0.779 F1-measure when a combination of contextual topic, and punctuation mark features are used. However, RNN demonstrates the lowest performance among all models in terms of recall and F1-measure while GRU shows 3.72\%, 17.79\% and 11.47\% higher precision, recall, and F1-measure than LSTM. In this study, SVM appears as the second highest model in terms of precision and F1-measure, obtains highest 0.679 recall when only textual features are used. On the other hand, NB exhibits lowest precision value 0.594 but performs better than RNN with recall and F1-measure. When textual features are used in combination with topic and punctuation mark, recall increases by 0.6\%, 2.5\%, and 11.92\%, and F1-measure increases by 0\%, 3.39\%, and 7\% for NB, SVM, and KNN models, respectively. Nevertheless, precision increases by 4.31\% and 1.49\% for SVM and KNN methods while decreases by 0.7\% in NB. \\

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM, KNN, and RNN  for detecting no boundaries in eCoaching text. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_not_boundary}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_not_boundary} summarizes the performance of NB, SVM, KNN, and RNN models for detecting ``not boundary'' class in eCoaching text. Performance is remarkably high in ``not boundary'' class compared to boundary detection. Similar to boundary detection, KNN consistently outperforms over all other methods but obtains 22.40\%, 50.07\%, and 36.26\% higher precision, recall, and F1-measure for contextual feature and 20.85\%, 33.96\%, and 27.47\% higher precision, recall, and F1-measure for combined features. However, RNN demonstrates the lowest performance and exhibits 0.981 and 0.983 precision with 0.986 and 0.987 F1-measure for LSTM and GRU, respectively. Impact of additional features is also consistent in ``not boundary'' classification. Results show that F1-measure increases by 0\%, and 0.1\% for SVM and KNN models although decreases by 0.1\% for NB. \\

\begin{table}[ht]
\centering
\caption{Weighted average performance of NB, SVM, KNN, and RNN for the segmentation of eCoaching text in detecting both ``boundary'' and ``not boundary'' classes. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
  \begin{tabular}{|l|l|l|l|p{0.15\linewidth}|p{0.15\linewidth}|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{contextual features only}} & \multicolumn{3}{|c|}{\textbf{contextual + punctuation marks (+ topics except RNN)}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-measure}\\ \hline

@@

Table~\ref{tab:result_weighted_avg} outlines the weighted average results of the experiment on the models for the segmentation of eCoaching text by classifying them into ``boundary'' and ``not boundary'' classes. Overall, KNN obtains the best performance with all metrics and RNN denotes the lowest performance among all methods. NB and SVM demonstrate moderate performace, obtain precision 0.975 and 0.981, recall 0.974 and 0.982, and F1-measure 0.975 and 0.983 when textual features are used. Effect of the additional features is also consistent as above, precision increases by 0.1\%, 0.2\%, and 0.3\%, recall increases by 0\%, 0.1\%, and 0.2\%, and F1-measure increases by 0\%, 0.2\%, and 0.3\%, for NB, SVM, and KNN methods, respectively, when combined features are used.\\

\section*{Discussion}
This study is the first large-scale efforts to evaluate the segmentation of eCoaching text. Experimental results indicate that KNN is the best model among all machine learning methods considered for this study. KNN achieved 0.986 F1-measure in overall, 0.779 and 0.993 F1-measures for detecting ``boundary'' and ``not boundary'', respectively. The robust performance of KNN provides the evidence that machine learning models are capable to learn information from the email exchange. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the ecoacing domain, and it can be successfully applied to other domain as well.

@@

In this paper, results are reported by each class to avoid confusion about the overall model performance. In addition, standard metrics: precision, recall, and F1-measure were used to eliminate doubt about the model performance because accuracy is misleading for imbalance dataset. AUC values are also outlined due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which was demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. NB shows the highest AUC values and achieved 0.978 for both cases while it provides lowest classification results except RNN. On the other hand, KNN and SVM exhibit 0.972 and 0.835 AUCs when only textual features are used; and 0.959 and 0.844 AUCs when a combination of textual, topic and punctuation mark features are used. Finally, RNN demonstrates lowest AUC values among all machine learning models, achieved AUC values 0.704 and 0.740 for LSTM and GRU, respectively.

@@

Punctuation mark plays an important role in segmentation boundary detection, and large numbers of errors were encountered by the false positive of boundary identification. Similarly, additional information is the common reason for the classified original segment into multiple segments. For example, [need help from April].

@@

As for future work, As our future work, we plan to evaluate our approach on other datasets involves in discourse analysis. We also plan to use a combination of machine learning and natural language approaches to improve model performance. For example, part-of-speech tagging and distance from the boundary of the sentence might significantly enhance the classification performance. The limitation of this study is that eCoaching text is collected from a single medical institute; formatting, style, and email segment can be different in other settings.

\section*{Conclusion}
Segmentation of eCoaching text is an integral part of developing an automated eCoaching intervention. Although several studies have done for the segmentation of clinical text into sections and sentences, none of them are used for the segmentation of text into a group of MI behavior in the setting of discourse analysis with email under the principle of motivational interviews. In this paper, we compared the performance of machine learning models for the task of segmentation of e-coaching text. We found out that k-nearest neighbor provides the best performance for the segmentation of text in terms of all performance metrics. Manual segmentation of e-coaching data is very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop effective behavioral interventions. Our proposed methods can help to identify individual text segments, which can be annotated directly with a classification model. This approach will help for developing fully automated eCoaching and also accelerate the pace of identifying effective communication strategies linked to healthy eating.
